---
title: "Workshop 4"
output: html_document
date: "2024-05-15"
---

##Workshop 3:

#installing packages: tidyverse
```{r}
install.packages("rlang")
library(rlang)
library(tidyverse)
```


#Example below:

```{r}
table1
#> # A tibble: 6 × 4
#>   country      year  cases population
#>   <chr>       <int>  <int>      <int>
#> 1 Afghanistan  1999    745   19987071
#> 2 Afghanistan  2000   2666   20595360
#> 3 Brazil       1999  37737  172006362
#> 4 Brazil       2000  80488  174504898
#> 5 China        1999 212258 1272915272
#> 6 China        2000 213766 1280428583
table2
#> # A tibble: 12 × 4
#>   country      year type           count
#>   <chr>       <int> <chr>          <int>
#> 1 Afghanistan  1999 cases            745
#> 2 Afghanistan  1999 population  19987071
#> 3 Afghanistan  2000 cases           2666
#> 4 Afghanistan  2000 population  20595360
#> 5 Brazil       1999 cases          37737
#> 6 Brazil       1999 population 172006362
#> # ... with 6 more rows
table3
#> # A tibble: 6 × 3
#>   country      year rate             
#> * <chr>       <int> <chr>            
#> 1 Afghanistan  1999 745/19987071     
#> 2 Afghanistan  2000 2666/20595360    
#> 3 Brazil       1999 37737/172006362  
#> 4 Brazil       2000 80488/174504898  
#> 5 China        1999 212258/1272915272
#> 6 China        2000 213766/1280428583


```

#How we make our dataset tidy is by following three interrelated rules. Each variable must have its own column.Each observation must have its own row.Each value must have its own cell.


##Like below, we apply the function mutate to compute the rate given two variables.

```{r}
# Compute rate per 10,000
table1 %>% 
  mutate(rate = cases / population * 10000)
#> # A tibble: 6 × 5
#>   country      year  cases population  rate
#>   <chr>       <int>  <int>      <int> <dbl>
#> 1 Afghanistan  1999    745   19987071 0.373
#> 2 Afghanistan  2000   2666   20595360 1.29 
#> 3 Brazil       1999  37737  172006362 2.19 
#> 4 Brazil       2000  80488  174504898 5.61 
#> 5 China        1999 212258 1272915272 1.67 
#> 6 China        2000 213766 1280428583 1.67

# Compute cases per year
table1 %>% 
  count(year, wt = cases)
#> # A tibble: 2 × 2
#>    year      n
#>   <int>  <int>
#> 1  1999 250740
#> 2  2000 296920

# Visualise changes over time
library(ggplot2)
ggplot(table1, aes(year, cases)) + 
  geom_line(aes(group = country), colour = "grey50") + 
  geom_point(aes(colour = country))
```
##Excersise 4.4.1:
#For each of the sample tables, describe what each observation and each column represents.Sketch out the processes you would use to calculate the rate for table2 and table3. You will need to perform four operations:Extract the number of TB cases per country per year.Extract the matching population per country per year.Divide cases by population, and multiply by 10,000. Store back in the appropriate place Hint: you haven’t yet learned the functions you need to actually perform these, but you can still think through the transformations!


##Pivoting data to make it tidy:

#The second step is to resolve one of the two common problems with untidy data. These are:One variable is spread across multiple columns.One observation is scattered across multiple rows

##To fix these we will pivot our data (i.e. move it around) into tidy form using two functions in tidyr: pivot_longer() to lengthen data and pivot_wider() to widen data. Let’s explore these functions a bit further.

##4.5.1: Lenghtening datasets

##Using  pivot_longer() splits the dataset by column, and reformats it into the tidy format of observations as rows, columns as variables and values as cell entries. The dataset is now longer (more rows, at left) than the ‘wide format’ data on the right. 

##Example:
```{r}
billboard
view(billboard)
#> # A tibble: 317 × 79
#>   artist       track               date.entered   wk1   wk2   wk3   wk4   wk5
#>   <chr>        <chr>               <date>       <dbl> <dbl> <dbl> <dbl> <dbl>
#> 1 2 Pac        Baby Don't Cry (Ke... 2000-02-26      87    82    72    77    87
#> 2 2Ge+her      The Hardest Part O... 2000-09-02      91    87    92    NA    NA
#> 3 3 Doors Down Kryptonite          2000-04-08      81    70    68    67    66
#> 4 3 Doors Down Loser               2000-10-21      76    76    72    69    67
#> 5 504 Boyz     Wobble Wobble       2000-04-15      57    34    25    17    17
#> 6 98^0         Give Me Just One N... 2000-08-19      51    39    34    26    26
#> # ℹ 311 more rows
#> # ℹ 71 more variables: wk6 <dbl>, wk7 <dbl>, wk8 <dbl>, wk9 <dbl>, ...


```
#In this dataset, each observation is a song. The first three columns (artist, track and date.entered) are variables that describe the song. Then we have 76 columns (wk1-wk76) that describe the rank of the song in each week. Here, the column names are one variable (the week) and the cell values are another (the rank). To tidy the billboard dataset we will use pivot_longer().

#This is the case where actual data values (wk1, wk2 etc.)  are in the column name, with each observation (row of data) being a song. We need to have the data in a format where each row is an observation (so-called long format).

```{r}
billboard |> 
  pivot_longer(
    cols = starts_with("wk"), 
    names_to = "week", 
    values_to = "rank"
  )
#> # A tibble: 24,092 × 5
#>    artist track                   date.entered week   rank
#>    <chr>  <chr>                   <date>       <chr> <dbl>
#>  1 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk1      87
#>  2 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk2      82
#>  3 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk3      72
#>  4 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk4      77
#>  5 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk5      87
#>  6 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk6      94
#>  #> # ℹ 24,082 more rows


```
#As you can see in the above code snippet, there are three key arguments to the pivot_longer() function: cols which specifies the columns you want to pivot (the ones that aren’t variables). Note: you could either use !c(artist, track, date.entered) OR starts_with('wk') because the cols argument uses the same syntax as select(). names_to which names the variable stored in the column names. We chose to name that variable week.values_to which names the variable stored in the cell values that we named rank

#Note that in the code "week" and "rank" are quoted because they are new variables that we are creating, they don’t exist yet in the data when we run the pivot_longer() call.

##Notice the NA values in the output above? It looks like “Baby Don’t Cry” by 2 Pac was only in the top 100 for 7 out of 76 weeks. Therefore, when we lengthened the data, the weeks where it wasn't on the charts became ‘NA.’ These NA’s were forced to exist because of the structure of the dataset not because they are actually unknown. Therefore, we can simply ask pivot_longer to remove them by adding the argument values_drop_na = TRUE as shown below:

```{r}
billboard |> 
  pivot_longer(
    cols = starts_with("wk"), 
    names_to = "week", 
    values_to = "rank",
    values_drop_na = TRUE
  )
#> # A tibble: 5,307 × 5
#>   artist track                   date.entered week   rank
#>   <chr>  <chr>                   <date>       <chr> <dbl>
#> 1 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk1      87
#> 2 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk2      82
#> 3 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk3      72
#> 4 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk4      77
#> 5 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk5      87
#> 6 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk6      94
#> # ℹ 5,301 more rows
```
##4.5.2: Pivoting longer

```{r}
df <- tribble(
  ~id,  ~bp1, ~bp2,
   "A",  100,  120,
   "B",  140,  115,
   "C",  120,  125
)

```


###However, we want our new (tidy) dataset to have three variables: id (which already exists)measurement (the column names) value (the cell values).To make this happen we need to pivot df longer:

```{r}
df |> 
  pivot_longer(
    cols = bp1:bp2,
    names_to = "measurement",
    values_to = "value"
  )
#> # A tibble: 6 × 3
#>   id    measurement value
#>   <chr> <chr>       <dbl>
#> 1 A     bp1           100
#> 2 A     bp2           120
#> 3 B     bp1           140
#> 4 B     bp2           115
#> 5 C     bp1           120
#> 6 C     bp2           125
```

#The left table represents what we originally typed when we created our dataset ‘df.’ The right table is what happened to the id when we used pivot_longer(). See how the data pivoted and repeated an id for each measurement? Additionally, the original column names in df (bp1 and bp2) now become values in a new variable, whose name is defined by the names_to argument and these values need to be repeated once for each row in the original dataset: And finally, the cell values also become a new variable with a name we defined by the values_to argument. These are unwound row by row:

##4.5.3: Widening datasets

###less common cases, we may need to widen a dataset rather than lengthen it. Widening is essentially the opposite of lengthening and we do so by using the function pivot_wider(). pivot_wider() allows us to handle an observation if it is scattered across multiple rows. Let’s visualize this in the image below:

```{r}
cms_patient_experience
#> # A tibble: 500 × 5
#>   org_pac_id org_nm                     measure_cd   measure_title   prf_rate
#>   <chr>      <chr>                      <chr>        <chr>              <dbl>
#> 1 0446157747 USC CARE MEDICAL GROUP INC CAHPS_GRP_1  CAHPS for MIPS...       63
#> 2 0446157747 USC CARE MEDICAL GROUP INC CAHPS_GRP_2  CAHPS for MIPS...       87
#> 3 0446157747 USC CARE MEDICAL GROUP INC CAHPS_GRP_3  CAHPS for MIPS...       86
#> 4 0446157747 USC CARE MEDICAL GROUP INC CAHPS_GRP_5  CAHPS for MIPS...       57
#> 5 0446157747 USC CARE MEDICAL GROUP INC CAHPS_GRP_8  CAHPS for MIPS...       85
#> 6 0446157747 USC CARE MEDICAL GROUP INC CAHPS_GRP_12 CAHPS for MIPS...       24
#> # ℹ 494 more rows

```
```{r}
cms_patient_experience |> 
  distinct(measure_cd, measure_title)
#> # A tibble: 6 × 2
#>   measure_cd   measure_title                                                 
#>   <chr>        <chr>                                                         
#> 1 CAHPS_GRP_1  CAHPS for MIPS SSM: Getting Timely Care, Appointments, and In...
#> 2 CAHPS_GRP_2  CAHPS for MIPS SSM: How Well Providers Communicate            
#> 3 CAHPS_GRP_3  CAHPS for MIPS SSM: Patient's Rating of Provider              
#> 4 CAHPS_GRP_5  CAHPS for MIPS SSM: Health Promotion and Education            
#> 5 CAHPS_GRP_8  CAHPS for MIPS SSM: Courteous and Helpful Office Staff        
#> 6 CAHPS_GRP_12 CAHPS for MIPS SSM: Stewardship of Patient Resources
```


###Neither of these columns will make particularly great variable names: measure_cd doesn’t hint at the meaning of the variable and measure_title is a long sentence containing spaces. We’ll use measure_cd as the source for our new column names for now, but in a real analysis you might want to create your own variable names that are both short and meaningful.pivot_wider() has the opposite interface to pivot_longer(): instead of choosing new column names, we need to provide the existing columns that define the values (values_from) and the column name (names_from):

```{r}
cms_patient_experience |> 
  pivot_wider(
    names_from = measure_cd,
    values_from = prf_rate
  )

```

###The above output doesn’t look quite right; we still seem to have multiple rows for each organization. That’s because, we also need to tell pivot_wider() which column or columns have values that uniquely identify each row; in this case those are the variables starting with "org":

```{r}
cms_patient_experience |> 
  pivot_wider(
    id_cols = starts_with("org"),
    names_from = measure_cd,
    values_from = prf_rate
  )
```

##4.5.4:Pivoting wider

```{r}
df <- tribble(
  ~id, ~measurement, ~value,
  "A",        "bp1",    100,
  "B",        "bp1",    140,
  "B",        "bp2",    115, 
  "A",        "bp2",    120,
  "A",        "bp3",    105
)
```

##We’ll take the names from the measurement column using the names_from() argument and the values from the value column using the values_from() argument:

```{r}
df |> 
  pivot_wider(
    names_from = measurement,
    values_from = value
  )

```


#To start the pivoting process, pivot_wider() needs to first figure out what will go in the rows and columns. The new column names will be the unique values of measurement.

```{r}
df |> 
  distinct(measurement) |> 
  pull()

```

###By default, the rows in the output are determined by all the variables that aren’t going into the new names or values. These are called the id_cols. Here there is only one column, but in general there can be any number.

```{r}
df |> 
  select(-measurement, -value) |> 
  distinct()
```














