---
title: "Workshop 4"
output: html_document
date: "2024-05-15"
---

##Workshop 3:

#installing packages: tidyverse
```{r}
install.packages("rlang")
library(rlang)
library(tidyverse)
```


#Example below:

```{r}
table1
#> # A tibble: 6 × 4
#>   country      year  cases population
#>   <chr>       <int>  <int>      <int>
#> 1 Afghanistan  1999    745   19987071
#> 2 Afghanistan  2000   2666   20595360
#> 3 Brazil       1999  37737  172006362
#> 4 Brazil       2000  80488  174504898
#> 5 China        1999 212258 1272915272
#> 6 China        2000 213766 1280428583
table2
#> # A tibble: 12 × 4
#>   country      year type           count
#>   <chr>       <int> <chr>          <int>
#> 1 Afghanistan  1999 cases            745
#> 2 Afghanistan  1999 population  19987071
#> 3 Afghanistan  2000 cases           2666
#> 4 Afghanistan  2000 population  20595360
#> 5 Brazil       1999 cases          37737
#> 6 Brazil       1999 population 172006362
#> # ... with 6 more rows
table3
#> # A tibble: 6 × 3
#>   country      year rate             
#> * <chr>       <int> <chr>            
#> 1 Afghanistan  1999 745/19987071     
#> 2 Afghanistan  2000 2666/20595360    
#> 3 Brazil       1999 37737/172006362  
#> 4 Brazil       2000 80488/174504898  
#> 5 China        1999 212258/1272915272
#> 6 China        2000 213766/1280428583


```

#How we make our dataset tidy is by following three interrelated rules. Each variable must have its own column.Each observation must have its own row.Each value must have its own cell.


##Like below, we apply the function mutate to compute the rate given two variables.

```{r}
# Compute rate per 10,000
table1 %>% 
  mutate(rate = cases / population * 10000)
#> # A tibble: 6 × 5
#>   country      year  cases population  rate
#>   <chr>       <int>  <int>      <int> <dbl>
#> 1 Afghanistan  1999    745   19987071 0.373
#> 2 Afghanistan  2000   2666   20595360 1.29 
#> 3 Brazil       1999  37737  172006362 2.19 
#> 4 Brazil       2000  80488  174504898 5.61 
#> 5 China        1999 212258 1272915272 1.67 
#> 6 China        2000 213766 1280428583 1.67

# Compute cases per year
table1 %>% 
  count(year, wt = cases)
#> # A tibble: 2 × 2
#>    year      n
#>   <int>  <int>
#> 1  1999 250740
#> 2  2000 296920

# Visualise changes over time
library(ggplot2)
ggplot(table1, aes(year, cases)) + 
  geom_line(aes(group = country), colour = "grey50") + 
  geom_point(aes(colour = country))
```
##Excersise 4.4.1:
#For each of the sample tables, describe what each observation and each column represents.Sketch out the processes you would use to calculate the rate for table2 and table3. You will need to perform four operations:Extract the number of TB cases per country per year.Extract the matching population per country per year.Divide cases by population, and multiply by 10,000. Store back in the appropriate place Hint: you haven’t yet learned the functions you need to actually perform these, but you can still think through the transformations!


##Pivoting data to make it tidy:

#The second step is to resolve one of the two common problems with untidy data. These are:One variable is spread across multiple columns.One observation is scattered across multiple rows

##To fix these we will pivot our data (i.e. move it around) into tidy form using two functions in tidyr: pivot_longer() to lengthen data and pivot_wider() to widen data. Let’s explore these functions a bit further.

##4.5.1: Lenghtening datasets

##Using  pivot_longer() splits the dataset by column, and reformats it into the tidy format of observations as rows, columns as variables and values as cell entries. The dataset is now longer (more rows, at left) than the ‘wide format’ data on the right. 

##Example:
```{r}
billboard
#> # A tibble: 317 × 79
#>   artist       track               date.entered   wk1   wk2   wk3   wk4   wk5
#>   <chr>        <chr>               <date>       <dbl> <dbl> <dbl> <dbl> <dbl>
#> 1 2 Pac        Baby Don't Cry (Ke... 2000-02-26      87    82    72    77    87
#> 2 2Ge+her      The Hardest Part O... 2000-09-02      91    87    92    NA    NA
#> 3 3 Doors Down Kryptonite          2000-04-08      81    70    68    67    66
#> 4 3 Doors Down Loser               2000-10-21      76    76    72    69    67
#> 5 504 Boyz     Wobble Wobble       2000-04-15      57    34    25    17    17
#> 6 98^0         Give Me Just One N... 2000-08-19      51    39    34    26    26
#> # ℹ 311 more rows
#> # ℹ 71 more variables: wk6 <dbl>, wk7 <dbl>, wk8 <dbl>, wk9 <dbl>, ...


```
#In this dataset, each observation is a song. The first three columns (artist, track and date.entered) are variables that describe the song. Then we have 76 columns (wk1-wk76) that describe the rank of the song in each week. Here, the column names are one variable (the week) and the cell values are another (the rank). To tidy the billboard dataset we will use pivot_longer().

#This is the case where actual data values (wk1, wk2 etc.)  are in the column name, with each observation (row of data) being a song. We need to have the data in a format where each row is an observation (so-called long format).

```{r}
billboard |> 
  pivot_longer(
    cols = starts_with("wk"), 
    names_to = "week", 
    values_to = "rank"
  )
#> # A tibble: 24,092 × 5
#>    artist track                   date.entered week   rank
#>    <chr>  <chr>                   <date>       <chr> <dbl>
#>  1 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk1      87
#>  2 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk2      82
#>  3 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk3      72
#>  4 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk4      77
#>  5 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk5      87
#>  6 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk6      94
#>  #> # ℹ 24,082 more rows


```
#As you can see in the above code snippet, there are three key arguments to the pivot_longer() function: cols which specifies the columns you want to pivot (the ones that aren’t variables). Note: you could either use !c(artist, track, date.entered) OR starts_with('wk') because the cols argument uses the same syntax as select(). names_to which names the variable stored in the column names. We chose to name that variable week.values_to which names the variable stored in the cell values that we named rank

#Note that in the code "week" and "rank" are quoted because they are new variables that we are creating, they don’t exist yet in the data when we run the pivot_longer() call.

